The role of cognateness in recognition of spoken words in a foreign language

### 2024-07-22 manuscript/manuscript.tex
New version of introduction

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The role of cognateness in native spoken word recognition},
  pdfauthor={Gonzalo Garcia-Castro; Serene Siow; Kim Plunkett; Nuria Sebastian-Galles},
  pdflang={en},
  pdfkeywords={cognate, spoken word
recognition, phonology, bilingualism, non-native speech
processing, lexical access},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Cognateness and native word recognition }
\title{The role of cognateness in native spoken word recognition}
\author{\textbf{Gonzalo
Garcia-Castro}~\orcidlink{0000-0002-8553-4209}\\Center for Brain and
Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\texttt{\href{mailto:gonzalo.garciadecastro@upf.edu}{gonzalo.garciadecastro@upf.edu}}\And\textbf{Serene
Siow}~\orcidlink{0000-0001-6482-2191}\\Department of Experimental
Psychology\\University of Oxford\\Oxford, OX2
6GG\\\texttt{\href{mailto:siow.serene@gmail.com}{siow.serene@gmail.com}}\And\textbf{Kim
Plunkett}~\orcidlink{0000-0003-0216-7480}\\Department of Experimental
Psychology\\University of Oxford\\Oxford, OX2
6GG\\\texttt{\href{mailto:kim.plunkett@psy.ox.ac.uk}{kim.plunkett@psy.ox.ac.uk}}\And\textbf{Nuria
Sebastian-Galles}~\orcidlink{0000-0001-6938-2498}\\Center for Brain and
Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\texttt{\href{mailto:nuria.sebastian@upf.edu}{nuria.sebastian@upf.edu}}}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Understanding spoken words in a non-native language is easier when their
translations into the listener's native language are phonologically
similar. Phonological similarity plays a more important role in the
language performance of low-proficiency bilinguals, compared to that of
high-proficiency bilinguals. This suggests that low-proficiency
bilinguals may points to the latter as relying more strongly on
cross-language word-to-word links (lexical route), while
high-proficiency bilinguals rely more on word-concept associations
(conceptual route) than on cross-language word-to-word links (lexical
route). Disentangling bilinguals' reliance on the conceptual and lexical
and conceptual routes during translation is challenging, given that both
sources of information are present during spoken word comprehension to
some extent in both high- and low-proficiency bilinguals. In this study,
we tested English and Spanish native speakers in a translation
elicitation task, in which they listened to words in an unfamiliar
language (Spanish or Catalan for English natives, and Catalan for
Spanish natives), and then had to translate them to their native
language. Given their lack of previous knowledge on the unfamiliar
tested language, phonological similarity between the presented words and
their correct translations was the only cue available for participants
to succeed in the task. This allowed us to explore the informativeness
of the lexical route for word translation, in the absence of information
from the conceptual route. Our results indicate that participants
benefited from the phonological similarity between translation
equivalents: the more similar the word they listened to and its
translational equivalent, the higher the probability of a correct
translation. These results point to cross-language word-to-word links as
providing rich information during word translation.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
cognate \sep spoken word
recognition \sep phonology \sep bilingualism \sep non-native speech
processing \sep 
lexical access

@@

Listening to non-native speech is costlier than listening to native
speech, even for highly proficient bilinguals, and especially in
acoustically adverse situations
(\citeproc{ref-lecumberri2010non}{Lecumberri et al., 2010};
\citeproc{ref-takata1990english}{Takata \& Nábělek, 1990}). One of the
sources of this increased difficulty is the partial mismatch between the
phonology of the native and the non-native languages: some acoustic
features embedded in the non-native speech signal do not overlap
perfectly with any phonemic category in the listener's native language.
For example, imagine a Spanish native with no previous familiarity with
any other language listening to French for the first time. This listener
may encounter the word /pɔʁt/ (\emph{porte}), translation of door in
French. The voiced uvular fricative consonant /ʁ/ and the open-mid back
rounded vowel /ɔ/ do not exist in Spanish as phonemic categories. These
discrepancies can nonetheless be perceived as allophonic variations the
native (Spanish) phonemes /ɾ/ and /o/
(\citeproc{ref-best2001discrimination}{Best et al., 2001}), leading to
the non-native speech signal engaging lexical processing mechanisms
(\citeproc{ref-weber2004lexical}{Weber \& Cutler, 2004}). Even if
unfamiliar with the word-form /pɔʁt/, the Spanish monolingual is likely
to activate the phonological representation of its Spanish translation
/pweɾta/ (\emph{puerta}), ultimately accessing the right semantic
representation. Although the phonetic mismatch between the non-native
word and its translation can have a noticeable toll on comprehension
(\citeproc{ref-cutler2004patterns}{Cutler et al., 2004}), the fact that
word recognition can take place at all in such circumstances illustrates
that non-native listeners---even low-proficiency ones---are rarely
completely naïve to the language they are listening to.

Imagine now that the same Spanish listener hears English for the first
time and encounters the word /dɔː/ (\emph{door}). This time,
phonological similarity is of little help: /pweɾta/ and /dɔː/ do not
share any phonemes. It would come as a surprise if the listener was able
to translate the English word successfully. Non-native listeners rely on
the phonological similarity between the word-forms presented in the
input and their translations. Luckily for them, most languages share
such kind of similarity. This is frequently due to their typological
closeness and/or socio-historical events involving the speakers of these
languages (e.g., migration, social contact). For example, Romance
languages such as Spanish and Catalan share many form-similar
translation equivalents (also known as \emph{cognates})
(\citeproc{ref-schepens2012distributions}{Schepens et al., 2012}), as in
the case of \emph{puerta} and \emph{porta} (door in Spanish and Catalan,
respectively). This type of cross-linguistic similarity has been
extremely informative for the study of bilingual lexical processing:
words that share high form-similarity with their translations are
processed faster and more accurately than words sharing lower
phonological similarity. This phenomenon has provided strong evidence in
favour of the language-non selective hypothesis of bilingual lexical
access, which states that bilinguals activate lexical representations in
both languages, even during monolingual situations. This facilitation
effect has been reported for comprehension
(\citeproc{ref-dijkstra2010cross}{Dijkstra et al., 2010};
\citeproc{ref-midgley2011effects}{Midgley et al., 2011};
\citeproc{ref-thierry2007brain}{Thierry \& Wu, 2007}), production
(\citeproc{ref-costa2000cognate}{Costa et al., 2000}), learning
(\citeproc{ref-de2000hard}{De Groot \& Keijzer, 2000};
\citeproc{ref-elias2022crosslanguage}{Elias \& Degani, 2022};
\citeproc{ref-lotto1998effects}{Lotto \& De Groot, 1998};
\citeproc{ref-valente2018does}{Valente et al., 2018}), and translation
(\citeproc{ref-christoffels2006memory}{Christoffels et al., 2006}). The
role that cross-linguistic phonological similarity might play in
monolingual lexical access during non-native speech comprehension is
unclear. How much phonological similarity is enough for a monolingual to
understand a word from a non-native language? How does phonological
similarity interact with other determinants of word recognition?

In the present study, we investigate the extent to which monolinguals
can exploit phonological similarity when listening to words in an
unfamiliar languages, in order to achieve comprehension, while taking
into account other lexically-relevant variables interact with the
lexical processing mechanisms involved. We used a translation
elicitation task in which participants listened to words from an
unfamiliar language, and then tried to guess their translation in their
native language. We will henceforth refer to the auditory-presented
words heard by participants on each trial as \emph{presented words}, and
the correct translation for the presented words as \emph{target
translations}. We explored listeners' reliance on phonological
similarity to succeed in the task by manipulating the amount of
phonological similarity between the presented words and their target
translations. Since participants were unfamiliar with the presented
language, they could only be able to use word-to-word connections to
succeed in the translation of words in an unfamiliar language. We
therefore predicted that, if the information provided by the
phonological form of the unfamiliar presented words is sufficient for
translation, participants' performance should increase when the
translation pairs are phonologically more similar.

The extent to which participants can rely on phonological similarity is
limited by other factors. In particular, the presence of phonological
neighbours (word-forms than differ in one phoneme) may interfere with
the activation and selection of the appropriate lexical representation
in the native language. Words with denser phonological neighbourhoods
are recognised more slowly and less accurately than words with sparser
phonological neighbourhoods (\citeproc{ref-dufour2003lexical}{Dufour \&
Peereman, 2003}; \citeproc{ref-goldinger1989priming}{Goldinger et al.,
1989}; \citeproc{ref-hamburger1996phonological}{Hamburger \& Slowiaczek,
1996}; \citeproc{ref-luce1990similarity}{Luce et al., 1990};
\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}). This is
specially true if such neighbours share higher phonological similarity
with the presented word, or are lexically more frequent than the target
translation (\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}).
The inhibitory effect of phonological neighbourhoods has also been
reported across languages. For instance, Weber \& Cutler
(\citeproc{ref-weber2004lexical}{2004}) observed that bilinguals
displayed lexical competition effects from words in their native
language when performing a word-referent matching task in their L2.

In the case of the present task, the potential effect of cross-language
phonological neighbourhood density is critical. The presence of
phonologically similar non-translation equivalents (e.g., false friends)
may act as distractor during lexical access, obstructing the selection
of the appropriate target translation in the listener's lexicon. Even if
the presented word and its target translation share high phonological
similarity, participants may still provide incorrect translations if the
presented word also shares high phonological similarity with words on
the target language other than its translation, specially if any of them
is more lexically frequent than the target translation. We hypothesised
that the size of the facilitatory role of phonological similarity would
be inversely proportional to the amount of higher-frequency phonological
neighbours of the presented word in the target language. For instance,
one could expect English participants to incorrectly translate the
Spanish word \emph{botón} as bottom instead of as its correct
translation \emph{button} (because the lexical frequency of
\emph{bottom} is higher than that of \emph{button} and also phonological
closer). To test this prediction, we developed a lexical
frequency-dependent measure of cross-language phonological neighbourhood
density, in which a neighbour is counted only if it is higher frequency
and is one phoneme appart from the presented word. If the phonological
neighbourhood density of the target translation affects participants'
performance negatively, this would suggest that non-native words trigger
competition between words in the native language during translation.

In Experiment 1, we collected data from two groups of British English
natives living in the United Kingdom. One group was presented with
Catalan words, the other with Spanish words. We examined the extent to
which participants were able to use the phonological similarity between
the presented word (in Catalan or Spanish) and its target translation to
provide accurate responses. In Experiment 2, we tested a group of
(European) Spanish natives in the same task, who were presented with
Catalan words. Catalan and Spanish are both Romance languages whose
close typological distance is reflected in the fact that they share many
cognates, where English is a Germanic language that shares considerably
fewer cognates with Catalan and Spanish. By testing participants
translating words from typologically close and distant languages, we
expected to widen the range of the phonological similarity scores of the
translation pairs involved in the experimental task, therefore allowing
us to explore potential cross-language differences in participant's
performance. Participants in Experiments 1 and 2 were surprisingly good
at translation some words, which had low phonological similarity with
their correct translation. In Experiment 3, we collected additional data
from a new group of British English natives, who were presented with
Catalan and Spanish words. In contrast with the procedure in Experiment
1, participants in Experiment 3 where asked, after providing a response
in each trial, to report whether they had previous knowledge of the
meaning of the presented word (e.g., some Spanish words are common
knowledge, even for monolingual speakers of English). Finally, we
present analyses on the joint data sets of Experiments 1 to 3.

@@

For each translation pair, we defined three predictors of interest: the
lexical frequency of the correct translation (\emph{Frequency}), the
phonological similarity between the presented (Catalan or Spanish) word
and their correct English translation (\emph{Cognateness}), and the
presented word's number of cross-language phonological neighbours
(\emph{CLPN}).

@@

We calculated \emph{Cognatenesss}, our main predictor of interest, by
computing the Levenshtein similarity between the X-SAMPA transcriptions
of each pair of translations using the \texttt{stringdist} R package
(\citeproc{ref-van2014stringdist}{van der Loo, 2014}). The Levenshtein
distance computes the edit distance between two character strings---in
this case, phoneme transcriptions---by counting the number of additions,
deletions, and substitutions necessary to make both strings identical
(\citeproc{ref-levenshtein1966binary}{Levenshtein, 1966}). We divided
this edit distance by the number of characters included in the longest
X-SAMPA transcription of the translation pair. This results in a
proportion score, in which values closer to zero correspond to lower
Levenshtein distances between phonological transcriptions (i.e., higher
similarity), and values closer to 1 correspond to higher Levenshtein
distances (i.e., lower similarity). This transformation also help
account for the fact that the Levenshtein distance tends to increase
with the length of the transcriptions. For interpretability, we
subtracted this proportion from one, so that values closer to one
correspond to higher similarity between phonological transcriptions, and
values closer to zero correspond to lower similarity between
phonological transcriptions. For example, the \emph{table}
(/teɪbəl/)-\emph{mesa} (/mesa/) translation pair had a 17\% similarity,
while the \emph{train} (\texttt{trEIn})-\emph{tren} (tɾEn) translation
pair had a 60\% similarity. Figure 1 summarises the lexical frequency,
phonological neighbourhood density and phonological overlap of the words
included in the Catalan and the Spanish lists.

@@

\toprule
\multicolumn{1}{l}{} & \multicolumn{2}{c}{Frequency} & \multicolumn{2}{c}{Cognateness} & \multicolumn{2}{c}{CLPN} \\ 
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7}
\multicolumn{1}{l}{} & Mean ± SD & Range & Mean ± SD & Range & Mean ± SD & Range \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & $5.88$ ± $0.57$ & $4.43$–$7.24$ & $0.13$ ± $0.18$ & $0.00$–$0.75$ & $0.19$ ± $0.79$ & $0$–$5$ \\ 
cat-ENG & $5.89$ ± $0.59$ & $4.43$–$7.27$ & $0.16$ ± $0.18$ & $0.00$–$0.67$ & $0.76$ ± $2.53$ & $0$–$15$ \\ 
cat-SPA & $5.79$ ± $0.63$ & $4.48$–$7.70$ & $0.38$ ± $0.26$ & $0.00$–$1.00$ & $0.87$ ± $2.08$ & $0$–$12$ \\ 
\midrule 
\midrule 
\textbf{Mean} & $5.85$ & — & $0.225$ & — & $0.606$ & — \\ 
\bottomrule

@@

After data collection, participants answers were manually coded into the
following categories: \emph{Correct}, \emph{Typo}, \emph{Wrong},
\emph{False friend}, \emph{Other}. Responses were coded as
\emph{Correct} if the provided string of characters was identical to the
orthographic form of the correct translation. Trials were coded as
\emph{Typo} if the participant provided a string of characters only one
edit distance (addition, deletion, or substitution) apart from the
orthographic form of the correct translation (e.g., ``pengiun'' instead
of ``penguin''), as long as the response did not correspond to a
distinct English word. Responses were coded as \emph{False friend} if
the participant provided a phonologically similar incorrect translation.
Responses were coded as \emph{Other} (see Data analysis section for more
details). Both \emph{Correct} and \emph{Typo} responses were considered
as correct, while \emph{Wrong} and \emph{False friend} responses were
considered as incorrect. \emph{Other} responses were excluded from data
analysis.

Trials in which participants took longer than 10 seconds to respond were
also excluded. Participants contributed a total of 9152 valid trials
(5694 in Catalan, 3458 in Spanish). The task took approximately 15
minutes to be completed.

@@

We modelled the probability of participants guessing the correct
translation of each presented word using a generalised multilevel
Bayesian regression model with a Bernoulli logit link distribution. We
included as fixed effects the intercept, the main effects of
\emph{Frequency}, \emph{Cognateness}, and \emph{CLPN}, and the two-way
interaction between \emph{Cognateness} and \emph{CLPN}. We also included
participant-level random intercepts and slopes for the the main effects
and the interaction. Eq. 1 shows a formal description of the model.

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The role of cognateness in recognition of spoken words in a foreign language},
  pdfauthor={Gonzalo Garcia-Castro; Serene Siow; Kim Plunkett; Nuria Sebastian-Galles},
  pdflang={en},
  pdfkeywords={cognate, spoken word
recognition, phonology, bilingualism, non-native speech
processing, lexical access},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Cognateness and recognition of foreign
words }
\title{The role of cognateness in recognition of spoken words in a
foreign language}
\author{\textbf{Gonzalo
Garcia-Castro}~\orcidlink{0000-0002-8553-4209}\\Center for Brain and
Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\texttt{\href{mailto:gonzalo.garciadecastro@upf.edu}{gonzalo.garciadecastro@upf.edu}}\And\textbf{Serene
Siow}~\orcidlink{0000-0001-6482-2191}\\Department of Experimental
Psychology\\University of Oxford\\Oxford, OX2
6GG\\\texttt{\href{mailto:siow.serene@gmail.com}{siow.serene@gmail.com}}\And\textbf{Kim
Plunkett}~\orcidlink{0000-0003-0216-7480}\\Department of Experimental
Psychology\\University of Oxford\\Oxford, OX2
6GG\\\texttt{\href{mailto:kim.plunkett@psy.ox.ac.uk}{kim.plunkett@psy.ox.ac.uk}}\And\textbf{Nuria
Sebastian-Galles}~\orcidlink{0000-0001-6938-2498}\\Center for Brain and
Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\texttt{\href{mailto:nuria.sebastian@upf.edu}{nuria.sebastian@upf.edu}}}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Understanding spoken words in a non-native language is easier when their
translations into the listener's native language are phonologically
similar. Phonological similarity plays a more important role in the
language performance of low-proficiency bilinguals, compared to that of
high-proficiency bilinguals. This suggests that low-proficiency
bilinguals may points to the latter as relying more strongly on
cross-language word-to-word links (lexical route), while
high-proficiency bilinguals rely more on word-concept associations
(conceptual route) than on cross-language word-to-word links (lexical
route). Disentangling bilinguals' reliance on the conceptual and lexical
and conceptual routes during translation is challenging, given that both
sources of information are present during spoken word comprehension to
some extent in both high- and low-proficiency bilinguals. In this study,
we tested English and Spanish native speakers in a translation
elicitation task, in which they listened to words in an unfamiliar
language (Spanish or Catalan for English natives, and Catalan for
Spanish natives), and then had to translate them to their native
language. Given their lack of previous knowledge on the unfamiliar
tested language, phonological similarity between the presented words and
their correct translations was the only cue available for participants
to succeed in the task. This allowed us to explore the informativeness
of the lexical route for word translation, in the absence of information
from the conceptual route. Our results indicate that participants
benefited from the phonological similarity between translation
equivalents: the more similar the word they listened to and its
translational equivalent, the higher the probability of a correct
translation. These results point to cross-language word-to-word links as
providing rich information during word translation.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
cognate \sep spoken word
recognition \sep phonology \sep bilingualism \sep non-native speech
processing \sep 
lexical access

@@

Learning a new language is a non-trivial task, a large part of which is
acquiring a new set of word-forms to correspond to existing semantic
representations in their lexicon. Imagine that a Spanish listener hears
English for the first time and encounters the word /\textipa{"dO:}/
(\emph{door}). Their knowledge of the phonological form in their native
language is of little help: /\textipa{"pweR.ta}/ and /\textipa{"dO:}/ do
not share any phonemes. It would come as a surprise if the listener was
able to translate the English word successfully without having first
learnt the word. However, there are some elements in language that may
help the learner in acquiring words in a new language. Many languages
share some degree of similarity. This may be due to typological
closeness and/or socio-historical events involving the speakers of these
languages (e.g., migration, social contact). For example, Romance
languages such as Spanish and Catalan share many form-similar
translation equivalents (also known as cognates)
(\citeproc{ref-schepens2012distributions}{Schepens et al., 2012}), as in
the case of \emph{puerta} and \emph{porta} (\emph{door} in Spanish and
Catalan, respectively). Given no prior knowledge of either language, a
Spanish native speaker is likely to be much more successful at correctly
translating Catalan \emph{porta} than English \emph{door}, due to the
phonological and orthographical similarity of the former to the
equivalent in their native language. Second language learners have been
found to learn cognates more easily than non-cognates
(\citeproc{ref-hall2002automatic}{Hall, 2002};
\citeproc{ref-otwinowska2019more}{Otwinowska \& Szewczyk, 2019}). This
type of cross-linguistic similarity has also been extremely informative
for the study of bilingual lexical processing: words that share high
form-similarity with their translations are processed faster and more
accurately than words sharing lower phonological similarity. This
phenomenon has provided strong evidence in favour of the language-non
selective hypothesis of bilingual lexical access, which states that
bilinguals activate lexical representations in both languages, even
during monolingual situations. This facilitation effect has been
reported for comprehension (\citeproc{ref-dijkstra2010cross}{Dijkstra et
al., 2010}; \citeproc{ref-midgley2011effects}{Midgley et al., 2011};
\citeproc{ref-thierry2007brain}{Thierry \& Wu, 2007}), production
(\citeproc{ref-costa2000cognate}{Costa et al., 2000}), learning
(\citeproc{ref-de2000hard}{De Groot \& Keijzer, 2000};
\citeproc{ref-elias2022crosslanguage}{Elias \& Degani, 2022};
\citeproc{ref-lotto1998effects}{Lotto \& De Groot, 1998};
\citeproc{ref-valente2018does}{Valente et al., 2018}), and translation
(\citeproc{ref-christoffels2006memory}{Christoffels et al., 2006}).

Event counting on this facilitation effect of form-similarity, listening
to non-native speech is costlier than listening to native speech, even
for highly proficient bilinguals, and especially in acoustically adverse
situations (\citeproc{ref-lecumberri2010non}{Lecumberri et al., 2010};
\citeproc{ref-takata1990english}{Takata \& Nábělek, 1990}). One of
potential source of this increased difficulty is the partial mismatch
between the phonology of the native and the non-native languages: some
acoustic features embedded in the non-native speech signal do not
overlap perfectly with any phonemic category in the listener's native
language. For example, imagine a Spanish native with no previous
familiarity with any other language listening to French for the first
time. This listener may encounter the word /\textipa{"pOKt}/
(\emph{porte}), translation of door in French. When comparing the
phonemic transcription to the transcription of the Spanish word
/\textipa{"pweR.ta}/ (\emph{puerta}), one may be led to infer that these
two words do not have much overlap. But one thing to note is that the
voiced uvular fricative consonant /\textipa{K}/ and the open-mid back
rounded vowel /\textipa{O}/, which do not exist in Spanish as phonemic
categories, can be perceived as allophonic variations the native
(Spanish) phonemes /\textipa{R}/ and /o/. A perceptual assimilation
model has been proposed to account for how speakers may integrate
non-native phonemes into their native phonemic categories based on
goodness-of-fit (\citeproc{ref-best1988examination}{Best et al., 1988},
\citeproc{ref-best2001discrimination}{2001}). Although phonetic mismatch
between the non-native word and its translation can have a noticeable
toll on comprehension Cutler et al.
(\citeproc{ref-cutler2004patterns}{2004}), the Spanish monolingual may
still be able to activate the phonological representation of
/\textipa{"pweR.ta}/ (\emph{puerta}) from hearing the word-form
/\textipa{"pOKt}/ (\emph{porte}), ultimately allowing them to access the
right semantic representation.

While phonological similarity across translation equivalents can help
with speech recognition, phonological similarity between words with
different meanings can hinder the recognition process. In word
retrieval, the presence of phonological neighbours (word-forms that
differ from each other in one phoneme) may interfere with the activation
and selection of the appropriate lexical representation in the native
language. Words with denser phonological neighbourhoods are recognised
more slowly and less accurately than words with sparser phonological
neighbourhoods (\citeproc{ref-dufour2003lexical}{Dufour \& Peereman,
2003}; \citeproc{ref-goldinger1989priming}{Goldinger et al., 1989};
\citeproc{ref-hamburger1996phonological}{Hamburger \& Slowiaczek, 1996};
\citeproc{ref-luce1990similarity}{Luce et al., 1990};
\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}). This is
especially true if such neighbours share higher phonological similarity
with the presented word, or are lexically more frequent than the target
translation (\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}).
The inhibitory effect of phonological neighbourhoods has also been
reported across languages. For instance, Weber \& Cutler
(\citeproc{ref-weber2004lexical}{2004}) observed that bilinguals
displayed lexical competition effects from words in their native
language when performing a word-referent matching task in their L2. The
assimilation of non-native phonemes into native categories
(\citeproc{ref-best1988examination}{Best et al., 1988},
\citeproc{ref-best2001discrimination}{2001}) was used to explain how
lexical competitors in the native language can be activated even when
the target words contain non-native phonemes. In the case of non-native
speech recognition, the presence of cross-linguistic pairs which are
phonologically similar but differ in meaning (e.g., false friends) may
act as distractors during lexical access, obstructing the selection of
the appropriate target translation in the listener's lexicon. For
instance, Otwinowska \& Szewczyk
(\citeproc{ref-otwinowska2019more}{2019}) reported that false friends
were disadvantaged relative to non-cognates by Polish second language
learners, in contrast to cognates which were known better. It is
therefore important to investigate the joint effect of both cognates and
false friends when investigating the effect cross-linguistic
phonological similarity has on word recognition in a foreign language.
This could shed light on available strategies and challenges associated
with the early stages of second language acquisition.

In the present study, we investigate the extent to which monolinguals
can exploit phonological similarity between translation equivalents when
listening to words in an unfamiliar language, and how it interacts with
competition from phonological neighbours in the native language. We used
a translation elicitation task in which participants listened to words
from an unfamiliar language, and then tried to guess their translation
in their native language. We will henceforth refer to the
auditory-presented words heard by participants on each trial as
presented words, and the correct translation for the presented words as
target translations. We explored listeners' reliance on phonological
similarity to succeed in the task by manipulating the amount of
phonological similarity between the presented words and their target
translations. Since participants were unfamiliar with the presented
language, they could only use phonological similarity between the
presented language and their native language to successfully translate
the words. We therefore predicted that participants' performance should
increase when the translation pairs are phonologically more similar. We
also predicted that there is a minimum threshold of phonological
similarity to be sufficient for translation.

We further investigated the effect of phonological competitors. Even if
the presented word and its target translation share high phonological
similarity, participants may still provide incorrect translations if the
presented word also shares high phonological similarity with other words
in the native language. The effect is predicted to be especially strong
if any of the competitors is more lexically frequent than the target
translation. We hypothesised that the size of the facilitatory role of
phonological similarity would be inversely proportional to the amount of
higher-frequency phonological neighbours of the presented word in the
target language. For instance, one could expect English participants to
incorrectly translate the Spanish word \emph{botón} as \emph{bottom}
instead of as its correct translation \emph{button}, due to combined
effect of the close phonological similarity between \emph{botón} and
\emph{bottom} along with the high lexical frequency of \emph{bottom}
relative to \emph{button}. To test this prediction, we developed a
lexical frequency-dependent measure of cross-language phonological
neighbourhood density, in which a neighbour is counted only if it is
higher frequency and is one phoneme apart from the presented word. If
the phonological neighbourhood density of the target translation affects
participants' performance negatively, this would suggest that
competitors in the native language affect recognition of non-native
words during recognition of foreign speech.We conducted a series of
three experiments to test these predictions.

In Experiment 1, we collected data from two groups of British English
natives living in the United Kingdom. One group was presented with
Catalan words, the other with Spanish words. We examined the extent to
which participants were able to use the phonological similarity between
the presented word (in Catalan or Spanish) and its target translation to
provide accurate responses. In Experiment 2, we tested a group of
(European) Spanish natives in the same task, who were presented with
Catalan words. Catalan and Spanish are both Romance languages whose
close typological distance is reflected in the fact that they share many
cognates, where English is a Germanic language that shares considerably
fewer cognates with Catalan and Spanish. By testing participants
translating words from typologically close or distant languages, we
expected to widen the range of the phonological similarity scores of the
translation pairs involved in the experimental task, therefore allowing
us to explore potential cross-language differences in participant's
performance. One unexpected finding was that participants in Experiments
1 and 2 were surprisingly good at translating a subset of words which
had low phonological similarity with their correct translation. We were
concerned that this may be caused by some prior knowledge of specific
words by our participants, as some Spanish words are commonly seen in
media or product labels, making them familiar even to monolingual
speakers of English. In Experiment 3, we collected additional data from
a new group of British English natives. The design was closely modelled
after Experiment 1, except that after providing their response in each
trial, participants reported whether they had previous knowledge of the
presented word. In the final section of this paper, we present analyses
on the joint data sets of Experiments 1 to 3.

@@

For each translation pair, we defined three predictors of interest: the
lexical frequency of the correct translation (\emph{Frequency}), the
phonological similarity between the presented (Catalan or Spanish) word
and their correct English translation (\emph{Similarity}), and the
presented word's number of cross-language phonological neighbours
(\emph{CLPN}).

@@

We calculated \emph{Similarity}, our main predictor of interest, by
computing the Levenshtein similarity between the X-SAMPA transcriptions
of each pair of translations using the \texttt{stringdist} R package
(\citeproc{ref-van2014stringdist}{van der Loo, 2014}). The Levenshtein
distance computes the edit distance between two character strings---in
this case, phoneme transcriptions---by counting the number of additions,
deletions, and substitutions necessary to make both strings identical
(\citeproc{ref-levenshtein1966binary}{Levenshtein, 1966}). We divided
this edit distance by the number of characters included in the longest
X-SAMPA transcription of the translation pair. This results in a
proportion score, in which values closer to zero correspond to lower
Levenshtein distances between phonological transcriptions (i.e., higher
similarity), and values closer to 1 correspond to higher Levenshtein
distances (i.e., lower similarity). This transformation also help
account for the fact that the Levenshtein distance tends to increase
with the length of the transcriptions. For interpretability, we
subtracted this proportion from one, so that values closer to one
correspond to higher similarity between phonological transcriptions, and
values closer to zero correspond to lower similarity between
phonological transcriptions. For example, the \emph{table}
(\texttt{teIb@l})-\emph{mesa} (\texttt{mesa}) translation pair had a
17\% similarity, while the \emph{train} (\texttt{trEIn})-\emph{tren}
(\texttt{tREn}) translation pair had a 60\% similarity.
Table~\ref{tbl-stimuli} summarises the lexical frequency, phonological
neighbourhood density and phonological overlap of the words included in
the Catalan and the Spanish lists.

@@

\toprule
\multicolumn{1}{l}{} & \multicolumn{2}{c}{Frequency} & \multicolumn{2}{c}{Similarity} & \multicolumn{2}{c}{CLPN} \\ 
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7}
\multicolumn{1}{l}{} & Mean ± SD & Range & Mean ± SD & Range & Mean ± SD & Range \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & $5.88$ ± $0.57$ & $4.43$–$7.24$ & $0.13$ ± $0.18$ & $0.00$–$0.75$ & $0.19$ ± $0.79$ & $0$–$5$ \\ 
cat-ENG & $5.89$ ± $0.59$ & $4.43$–$7.27$ & $0.16$ ± $0.18$ & $0.00$–$0.67$ & $0.76$ ± $2.53$ & $0$–$15$ \\ 
cat-SPA & $5.79$ ± $0.63$ & $4.48$–$7.70$ & $0.38$ ± $0.26$ & $0.00$–$1.00$ & $0.87$ ± $2.08$ & $0$–$12$ \\ 
\midrule 
\midrule 
\textbf{Mean} & $5.85$ & — & $0.225$ & — & $0.606$ & — \\ 
\bottomrule

@@

After data collection, participants answers were manually coded into the
following categories: \emph{Correct}, \emph{Typo}, \emph{Wrong},
\emph{False friend}, \emph{Other}. Responses were coded as
\emph{Correct} if the provided string of characters was identical to the
orthographic form of the correct translation. Trials were coded as
\emph{Typo} if the participant provided a string of characters only one
edit distance (addition, deletion, or substitution) apart from the
orthographic form of the correct translation (e.g., ``pengiun'' instead
of ``penguin''), as long as the response did not correspond to a
distinct English word. Responses were coded as \emph{False friend} if
the participant provided a phonologically similar incorrect translation.
Responses were coded as \emph{Other} (see Data analysis section for more
details). Both \emph{Correct} and \emph{Typo} responses were considered
as correct, while \emph{Wrong} and \emph{False friend} responses were
considered as incorrect. \emph{Other} responses were excluded from data
analysis. Trials in which participants took longer than 10 seconds to
respond were also excluded. Participants contributed a total of 9152
valid trials (5,694 in Catalan, 3,458 in Spanish). The task took
approximately 15 minutes to be completed.

@@

We modelled the probability of participants guessing the correct
translation of each presented word using a generalised multilevel
Bayesian regression model with a Bernoulli logit link distribution. We
included as fixed effects the intercept, the main effects of
\emph{Frequency}, \emph{Similarity}, and \emph{CLPN}, and the two-way
interaction between \emph{Similarity} and \emph{CLPN}. We also included
participant-level random intercepts and slopes for the the main effects
and the interaction. Eq. 1 shows a formal description of the model.

\[
\begin{aligned}
&\textbf{Likelihood}  \\
y_{i} \sim & \text{Bernoulli}(p_{i}) \\ \\
&\textbf{Parameters}  \\
\text{Logit}(p_{i}) = &  \beta_{0[p,w]} + \beta_{1[p]} \text{Frequency}_{i} + \beta_{2[p]} \text{PTHN}_i + \beta_{3[p]} \text{Similarity}_i + \beta_{4[p]} (\text{PTHN}_i \times \text{Similarity}_i) \\
\beta_{0-6[p,w]} \sim & \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p \text{ in 1, ..., } P \text{ and  word } w \text{ in 1, ..., } W \\
\beta_{1-6[p]} \sim &  \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p \text{ in 1, ..., } P \\ \\
&\textbf{Prior}  \\
\mu_{\beta_{p,w}}  \sim &  \mathcal{N}(0, 0.1) \\
\sigma_{\beta_{p}},  \sigma_{\beta_{w}} \sim & \text{HalfCauchy}(0, 0.1) \\
\rho_{p}, \rho_{w} \sim & \text{LKJCorr}(8) \\
\end{aligned}
\]
